# A Tour of R

## It's all about functions

Functions are often described as boxes. These boxes take inputs and create outputs. I think understanding what we are able to do with functions will help you understand how to solve the problems you are going to face. What we really care about is the ability to port around complex tools that make our tasks simpler. For example, functions will allow us to do the following and much more

* Access/build APIs and databases, store data locally or remote.
* Build models and make predictions based on the data we collect.
* Edit and manipulate our data into new, more useful forms
* Display our data on the internet in the form on web based applications. 

### Characteristics of functions

Functions can be named or unnamed. Unnamed functions are called anonymous functions. Anonymous functions are generally used when you have a simple but unique task that is applied to multiple objects sequentially, but which also is unlikely to be repeaded anywhere else. Let's look at a few functions to get acquainted with R how they work. 

Functions have an input and an output. The input is whatever is contained within the parenthesis `()` and the output will be displayed below the code. We will talk more about functions later, so don't try to get too caught up in the details here. Programming is something I want you to soak in, not memorize. The `c` function creates something called a vector, which is a $1$ dimensional matrix or table. Vectors are the singular columns or rows in a table. They do not mix types. The types you will work with are primarily logical, integer, double, character, list, NULL, closure (function).

There are two functions that we should get familiar with first, but let's go with the one I think you will use more of the two, though both `list` and `c` will be functions you use daily. 

```{r}
c(integer = 1L, double = 1, bool = TRUE, character = 'a')
```
```{r}
list(integer = 1L, double = 1, bool = TRUE, character = 'a')
```

Get used to looking at data like this. We're not in Excel anymore and sometimes it makes people feel funny about not being able to "see" their data. When I say see, I simply mean is that look at the data below. This is a two dimensional table called a dataframe or tibble. A dataframe is a special form of a list, that requires each element be atomic (of one type) and of equal length. We can see this is true by looking at the output below. 

```{r, echo=FALSE}
dplyr::glimpse(nycflights13::flights)
```

There are $19*336,776$ cells within this table because there are $19$ vectors (columns) of length $336,776$. Each element in the vector is one instance of a string, a number, a date, etc. And each type has functions that can operate on it, in a repeatable way. This is why each column (vector) is of one type, either `int`, `dbl`, `chr`, or `dttm`. There are other types in R, but we will be sticking to these for now. `int` is a special form of number called an integer - 1, 2, -1231, etc. `dbl` stands for double, a number which can contain decimal values i.e, 1.3. Not everything needs to be double, because an integer takes less memory in your computer to store. Don't get too caught up on whether or not something should be an integer or double and generally speaking you wont consciously make a choice.  


```{r}
library(purrr)
library(cli)
iwalk(c(number = 1, bool = TRUE, character = 'a'), function(item, name) cli_alert_info('{name} is a {typeof(item)}'))
```

```{r}
iwalk(list(number = 1, bool = TRUE, character = 'a'), function(item, name) cli_alert_info('{name} is a {typeof(item)}'))
```

The following unnamed function, a special case called an anonomous function or lambda function, $(x, y)\rightarrow x(x + y)$ has two critical components. Don't be afraid of the right arrow. We just say that $x$ and $y$ map _onto_ whatever expression is on the right hand side. 

1. The function has any number of inputs, in the case above there are two: $(x,y)$
2. The function has a body which defines the output, $x(x + y)$ 
3. We can express evaluation of the function by following the function with inputs:
  * $((x, y)\rightarrow x(x + y))(1, 2)$ which evaluates to $(1, 2) \rightarrow 1 * (1 + 2) = 3$


When we write functions in R we will use both named and unnamed functions. But lets start with the first example. This function is an anonymous function. We haven't given it a name yet. But we can call it all the same.

```{r}
function(x, y) x * (x + y)
```

```{r}
# Wrapping function in () means evaluate now
(function(x, y) x * (x + y))(1, 2)
```

Now wouldn't it be nice if we could take this function with us? Let's give it a name and update our notation $xy\_calc(x, y)\ \rightarrow x(x + y)$

```{r}
library(cli)

xy_calc <- function(x, y) {
  x_name <- deparse1(substitute(x))
  y_name <- deparse1(substitute(y))
  cli_alert_info('{x_name} is {x} and {y_name} is {y}')
  result <- x * (x + y)
  cli_alert_success('result is {result}')
  result
}

new_calc <- function(a, b) xy_calc(xy_calc(a, a), xy_calc(b, b))
```

Now we can describe silly things like this $newcalc(x, y) \rightarrow xy\_calc(xy\_calc(x, x), xy\_calc(y,y)))$

```{r}
new_calc(1, 2)
```


If we were to write by hand the expression we created, it would be



```{r}
library(cli)

# Same as f(x) = 2x + 3
f <-  function(x) {
  x <- x * x
  x
}

g <- function(x=NULL, y=NULL) {
  result <- x + min(x, y)
  result
}

print(f(g(3, 4)))
```

We now have a way of describing inputs and output a little more clearly. Instead of writing, $(3 + min(3, 4)) * (3 + min(3, 4))$ we can write $f(g(3, 4))$ or try new creations like $z(x, y)=f(g(f(x), f(y)))$ so $z(1, 2)=f(g(f(1), f(2)))=f(g(1, 4))=f(2)=4$.\
\
Now just take this idea about functions and expand your definition of inputs and outputs to be any number, none or many, and of any type that R supports - character, numeric, date/time, data.frame or list - all of which we'll cover.

## Solve a Problem in R

Let's solve a problem using R. Suppose we have a friend that is interested in the current trend regarding COVID-19 cases. The first thing we will probably do is try to figure out an efficient and reliable way for importing Covid-19 data into our R session. Conveniently, the [`COVID19`](https://covid19datahub.io/articles/r.html) package allows us to pull the latest data without any hard work and consists of one function - `covid19`.

### Installing Packages

In your RStudio console, you can write the following to install the `COVID19` package using the `install.packages` function. If you are interested in learning more about this function, you can write `?install.packages` in your console and the documentation for the function will appear.

```{r, eval=FALSE}
# For help menu, uncomment next line
# ?install.packages 

# If the package is not yet installed, you can install it by passing 
# a string with the package name to the `install.packages` function
install.packages(pkgs = c('COVID19'))
```

### Available Packages on CRAN

For a full list of what packages are available through the `install.packages` function, please check out the [Contributed Packages](https://cran.r-project.org/web/packages/index.html) page at CRAN or scrape it yourself.\

```{r, cache=TRUE}
library(rvest)
cran_packages <- 'https://cran.r-project.org/web/packages/available_packages_by_date.html'
package_data <- html_table(html_element(read_html(cran_packages), 'table'))
print(package_data)
n_packages <- length(unique(package_data$Package))
cli_alert_info('There are {n_packages} packages on CRAN')
```

### Using Functions to Solve a Problem

The code below consists of three different functions. The first two are `library` and `covid19`, but the third is hidden - it's actually the arrow, `<-` if you execute `` `<-`(a, 1) `` the output of the function actually creates the variable `a` within your session! *Functions* are spaces for stuff to happen. Functions help us make common procedures repeatable. By creating a function with a particular name and inputs, we can get some sort of useful (or not useful, the world's your oyster) output.

In this case, `library` loads packages from a folder in the R environment called library. You can see which ones your R environment knows about by running the function `.libPaths()`. The dot in front of `.libPaths()` just means that the author intended it to be hidden, which doesn't really mean much to us. When you run `install.packages` that code is at a path in the `.libPaths()` output.

`covid19` is a function from the `COVID19` package, and would only be available after executing `library(COVID19)` or if `library(COVID19)` is omitted, by pulling it from the package namespace directly by preceding the function with the package name and two colons like so: `COVID19::covid19`. Generally speaking, you simply use `library` because it reduces the amount of text on the page.

```{r, warning=FALSE, message=FALSE, cache=FALSE}
library(tidyverse)
library(purrr)
library(COVID19)
```

```{r, warning=FALSE, message=FALSE, cache=TRUE}
covid_data <- covid19(
    country = 'United States', 
    start = '2021-01-01', 
    end = "2021-11-21",
    verbose = FALSE, 
    level = 2
)
glimpse(covid_data)
```

Let's look at what happened - we passed a few inputs and received a dataframe. A dataframe is a list with the requirement that all elements of the list are atomic vectors of equal length. Let's look at what that means.

```{r}
map_chr(covid_data, typeof)
```

When you have a list of things, you can apply a function to each item in the list. So in the list above, we have 47 atomic vectors. What does that mean? An atomic vector is like a list, but it has to contain the same thing in each cell.

[How do I know if a function is vectorized](https://stackoverflow.com/questions/58568392/how-do-i-know-a-function-or-an-operation-in-r-is-vectorized#:~:text=To%20identify%20if%20an%20R%20object%20is%20a%20vector%20%2C%20I,a%20vector%20or%20False%20otherwise.)\
[Vectorization in R](https://www.noamross.net/archives/2014-04-16-vectorization-in-r-why/)

```{r}
vector_example <- c(1, 'a', TRUE)
list_example <- list(1, 'a', TRUE)

map_chr(vector_example, typeof)
map_chr(list_example, typeof)
```

With the knowledge of vectors and lists, what can we do? Well, the first thing I notice is that some of the vectors are completely `NA`. Let's check the number of NA values in each vector.

```{r}
all_na <- function(item) {
  sum(is.na(item))==length(item) 
}
covid_data <- discard(covid_data, all_na)
head(covid_data)
```
