---
title: 'A Handbook on R'
author: "Freddy Ray Drennan"
date: "`r Sys.Date()`"
output: pdf_document
documentclass: book
bibliography:
- book.bib
- packages.bib
biblio-style: apalike
link-citations: yes
description: A labor of love about R
site: bookdown::bookdown_site
---

# Introduction

## Initial Setup

[Book Outline](https://hackmd.io/vGRGEPo8QQyiG8gecWv71g)

-   [Install R](https://cran.r-project.org/)

-   [Install R Studio](https://www.rstudio.com/products/rstudio/download/)

-   [Windows Only: Install RTools](https://cran.r-project.org/bin/windows/Rtools/)

    -   When installed, run in the RStudio Console: `write('PATH="${RTOOLS40_HOME}\\usr\\bin;${PATH}"', file = "~/.Renviron", append = TRUE)`

-   [Windows Only: Install WSL2](https://www.omgubuntu.co.uk/how-to-install-wsl2-on-windows-10)

    -   Computer should be completely updated before install.

-   [Install Git](https://git-scm.com/downloads)

-   [Create Github Account](https://github.com/)

-   [Fork r-handbook](https://github.com/fdrennan/r-handbook)

-   [Install Docker and Docker Compose](https://docs.docker.com/get-docker/)

-   [Create AWS Account](https://aws.amazon.com/)

    -   Billing will be discussed in the course, but don't expect to pay much - i.e., 10-20 dollars a month for high course activity.
    -   Remember to `stop` EC2 servers when we begin using them. AWS is polite about your first few refund requests.

-   [Create Reddit Account](reddit.com)

    -   [Follow Instructions here](https://towardsdatascience.com/how-to-use-the-reddit-api-in-python-5e05ddfd1e5c)

Make sure you install the [`tidyverse`](https://www.tidyverse.org/) packages. Update to renv later.

## A Dose of Theory

I struggle to determine the best way to start a book like this. But, as someone  who has read one or two programming books, I can't help but get tired of the "Hello World" introduction. If you don't know what I am talking about, don't worry, you will but I won't bother you with that here. Instead, I want to talk about boring lists. 

A list is a container for anything that exists in the R world. Lists can be named or unnamed.


```{r}
list_with_values <- list(
    first_value_name = 1, 
    hello = mtcars, 
    model = lm(cyl ~ mpg, mtcars)
)
```


[How do I know if a function is vectorized](https://stackoverflow.com/questions/58568392/how-do-i-know-a-function-or-an-operation-in-r-is-vectorized#:~:text=To%20identify%20if%20an%20R%20object%20is%20a%20vector%20%2C%20I,a%20vector%20or%20False%20otherwise.)




```{r eval=FALSE}
install.packages('tidyverse')
```
